---
title: 'IMT 573 Assignment 1 - Exploring data'
author: "Srinivas Murri"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    highlight: tango
  html_document:
    df_print: paged
urlcolor: blue
---

## Install required packages and Libraries

```{r Installing Required Packages}
## installing package nycflights13 to get NYC Flights 2013 data.
##install.packages("nycflights13")
## importing the packages in to current session , Load some helpful libraries
library(nycflights13)

library(tidyverse) 

library(tinytex) ##to knit the file to PDF format
```

## Load data in to Data Frame

```{r Load 2013 NYC flights Data}
data(flights,package="nycflights13")
# Note: this is correct on the server
# Note: can read compressed files directly
```

Loaded the data in to the data frame "flights"

## Head Sample data
```{r Sample Top Records}
## here your code for inspecting data
## you may add more code chunks as needed
# Sample Data
head(flights)
```

## Tail Sample data
```{r Sample Bottom Records}
## here your code for inspecting data
## you may add more code chunks as needed
# Sample Data
tail(flights)
```

## Inspect data
```{r Data details}
## size of the data frame
dim(flights)
## column names
colnames(flights)
## Count of Rows in each column with NA rows
colSums(is.na(flights))
## Column wise datatype
sapply(flights, typeof)

```
The below variables have values NA, Corresponding NA count:

dep_time	8255
dep_delay	8255
arr_time	8713
arr_delay	9430
tailnum	2512
air_time	9430


### Summary stats of the dataset variables
```{r summary}
summary(flights,4)
```


```{r histograms of delays}
hist((flights %>% filter(arr_delay != 0) %>% select(arr_delay))$arr_delay,
main="Arrival Delay",
xlab="Arrival",
col="darkmagenta",
freq= FALSE,
breaks=200,
xlim=c(-100,100),  ylim=c(0, .02)
)

## Arrival delay is Normal

hist((flights %>% filter(dep_delay != 0) %>% select(dep_delay))$dep_delay,
main="Dep Delay",
xlab="Departure",
col="darkmagenta",
freq= FALSE,
breaks=200,
xlim=c(-100,100),  ylim=c(0, .02)
)
  
hist((flights$dep_time+flights$arr_time)/200 ,
main="Departure Hours",
xlab="Dep Time",
col="Yellow",
freq= FALSE,
breaks=50,
xlim=c(0,25),  ylim=c(0, .1)
)
## Airport is busy from 6:00 AM to 11:00 PM
```

## Listed Findings

### Findings

1. Arrival delay is nearly Normal, Most of the Arrival delay is towards negative also. Arrivals are earlier than scheduled
2. Departure delay is Skewed , Many Dep delays are also negative means departured before scheduled. However, most of the departure delays are spread towards postive and number of flights delayed lessens with duration of delay
3. Number of flights between 6 Am and 11 PM are spread even accross the day and the distribution is nearly Uniform



```{r Arrival delay vs Dep delay}

##DEP Delay vs ARR Delay
ggplot(flights %>% filter(dep_delay <1000), 
       aes(x = dep_delay/60, y = arr_delay/60)) +
  geom_point( color = "magenta" , alpha = 0.1) +
  geom_smooth( color = "blue" , alpha = 0.5 , size=0.2) +
  labs(x ="Dep Delay in Hrs",
       y =  "Arr Delay in Hrs" ,
       title = "Dep vs Arr Plot")
## More the departure delay, more the Arrival delay
```

### Findings
We have negative delays also ,but most of the delays are positive and less. There are occassions where the delay is beyond 10 hours also.
The Arrival delay is directly influenced by Departure delay and they are linearly dependent . The more the departure delay, the more the arrival delay


```{r monthly dep delays}
## created a data frame monthly maen flight dep delay 
monfltdel <- flights %>% 
filter(dep_delay > 0) %>%
group_by(month) %>%
summarize(delay=mean(dep_delay) , delaycnt=length(dep_delay)) %>%
arrange(month)

head(monfltdel)

ggplot(monfltdel, aes(x = factor(month), y = delay ))+
  stat_smooth( color = "red" , alpha = 1) +
  geom_point(size = 4, color = "red")+
  geom_bar(size = 0.1, color = "grey", fill = "steelblue",
           stat = "identity")+
  labs(x ="Month",
       y =  "Dep delay mean" ,
       title = "Month wise dep delay")
```

### Findings
From the above Bar chart , we can see most of the delays are in the months of June and July . Less delays in November. As dep delay and arr delay and linearly dependent. we can say arrival delays are also more in June and July


```{r Time wise delay}
timfltdel <- flights %>% 
filter(dep_delay > 0) %>%
group_by(time = round(sched_dep_time/100)) %>%
summarize(delay=mean(dep_delay) , delaycnt=length(dep_delay))  %>% arrange(time)

head(timfltdel)

## number of flight delays and more between 4 PM and 10 PM
ggplot(timfltdel, )+
  geom_bar(mapping = (aes(as.character(x = time), y = scale(delay,1) )),
           size = 0.2 , color = "red" ,fill = "steelblue", stat = "identity")+
    geom_point(mapping = (aes(as.character(x = time), y = scale(delaycnt,1)  )),
               size = 4 , color = "red" ,fill = "red", stat = "identity")+
    labs(x ="Hour Time",
       y =  "Dep delay mean" ,
       title = "Hourly dep delay")
  

## number of delay vs mean of delay .. more the number of flights delayed, more the average delay  time
ggplot(timfltdel )+
  geom_point(mapping = aes(x = delaycnt, y = delay),size = 2 , color = "darkblue")+
  geom_smooth(mapping = aes(x = delaycnt, y = delay),size = 0.2 , color = "red")+
  labs(x ="Flight delayed count",
       y =  "Dep delay mean" ,
       title = "Busy vs Delay")

```

### Findings

* From the above Bar chart , we can see most of the delays are between 4Pm and 10 PM.
* From Scatter plot we can say more the flights more the delays. not straight dependency but from smoothing we can see there is some dependency on delay with Busy


```{r data frames}
## carrier wise dep delay
carifltdd <- flights %>% 
filter(dep_delay > 0) %>%
group_by(carrier) %>%
summarize(ddelay=mean(dep_delay) )  %>% arrange(desc(ddelay))

head(carifltdd)

## carrier wise arr delay
carifltad <- flights %>% 
filter(arr_delay > 0) %>%
group_by(carrier) %>%
summarize(adelay=mean(arr_delay) )  %>% arrange(desc(adelay))

head(carifltad)

## destination wise arr delay
destarrdel <- flights %>% 
filter(arr_delay > 0) %>%
group_by(dest) %>%
summarize(adelay=mean(arr_delay) )  %>% arrange(desc(adelay))

head(destarrdel)

## destination wise dep delay
destdepdel <- flights %>% 
filter(dep_delay > 0) %>%
group_by(dest) %>%
summarize(ddelay=mean(dep_delay) )  %>% arrange(desc(ddelay))

head(destdepdel)

########################################################## PLOTS of above data sets

## carrier wise dep delay
ggplot( )+
  geom_bar(carifltdd ,mapping = (aes(   reorder(x=carrier,-ddelay) , y = ddelay  )),
           size = 0.1 , color = "white" ,fill = "green", stat = "identity")+
  labs(x ="Carrier",
       y =  "Dep delay mean" ,
       title = "Carrier wise dep delay")

## carrier wise arrival delay
ggplot( )+
  geom_bar(carifltad ,mapping = (aes(   reorder(x=carrier,-adelay) , y = adelay  )),
           size = 0.1 , color = "white" ,fill = "red", stat = "identity")+
  labs(x ="Carrier",
       y =  "Arr delay mean" ,
       title = "Carrier wise Arr delay")

## Origin wise arrival delay
ggplot( )+
  geom_bar(destarrdel[1:10,] ,mapping = (aes(   reorder(x=dest,-adelay) , y = adelay  )),
           size = 0.1 , color = "white" ,fill = "magenta", stat = "identity")+
  labs(x ="Destination",
       y =  "Arr delay mean" ,
       title = "Dest wise Arr delay")

## dest wise dep delay
ggplot( )+
  geom_bar(destdepdel[1:10,] ,mapping = (aes(   reorder(x=dest,-ddelay) , y = ddelay  )),
           size = 0.1 , color = "white" ,fill = "blue", stat = "identity")+
  labs(x ="Destination",
       y =  "Dep delay mean" ,
       title = "Dest wise Dep delay")

```

### Findings

* The OO , YV, EV, 9E , F9 are the carriers with more Departure and Arrival delays.
* TVC, TUL, BHM , TYS are the destinations with more Departure and Arrival delays.

